Automatically proving equivalence by type-safe reflection

### 2016-01-12 paper-JFP-DTP/content/5-discussion.tex
paper continued and corrected. Still roughly 2 pages to write

Without our type-safe reflection mechanism, the traditional way to go for this problem would be to have functions doing only computations, and some external lemmas about them. It would start with the definition of the function $reify\ :\ Expr\ \rightarrow\ c$. As we did in our approach, there is still a function for reflecting terms $reflect\ :\ c\ \rightarrow\ Expr$. Because $reflect$ is a function based on syntax,  the actual way to define depends on the facilities of the proof assistant / programming language used, but there need to be a way to write functions based on syntax. We need to prove the correctness of these two functions, and it turns out that one lemma is enough for completely specifying the expected behaviour of these two functions at the same time : $reflect\_and\_reify\_correct\ :\ \forall\ x:c,\ reify\ (reflect\ x)\ =\ x$.

Then comes the normalisation function : $normalize\ :\ Expr\ \rightarrow\ Expr$. Because this function is weekly typed compared to our approach (we no longer have an index giving a guarantee about the concrete values), we need to provide a lemma of correctness about it after its definition. This lemma of correctness must say that the interpretation of the original (reflected) term is equal to the interpretation of the normalised (reflected) term. $normalize\_correct\ :\ \forall\ e:Expr,\ reify\ (normalize\ e)\ =\ reify\ e$. We have avoided the need of such a --complicated-- proof in our correct by construction approach enabled by our type-safe reflection mechanism.

Now, similarly to what we've done, we need a syntactical equality test, checking whether two reflected expressions are syntactically equal. The difference here is that, as often with this kind of "traditional" approach, we produce something that belongs to the world of computations, usually a (pretty uninformative) boolean.
$beq_{Expr}\ :\ Expr\ \rightarrow\ Expr\ \rightarrow\ bool$. Because this boolean on its own is pretty uninformative, we now need a proof of correctness for this function as well, which says that if this function decides that two given terms are syntactically equal, then their interpretation should also be (syntactically) equal. $beq_{Expr}\_correct\ :\ \forall (e1\ e2:Expr), beq_{Expr}\ e1\ e2\ =\ true\ \rightarrow\ reify\ e1\ =\ reify\ e2$. We did not have to do this proof either with our approach.

We now build the kernel of the tactics, which is the following decision procedure. 
\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
decideEq : c -> c -> bool 
decideEq x y =
      let xRefl = reflect x in 
      let yRefl = reflect y in
      beq_Expr (normalise xRefl) (normalise yRefl)
\end{verbatim}
\end{center}
\caption{Decision procedure of the tactic with a "traditional" approach}
\figrule
\end{figure}
This function on its own is not going to be enough. We need to add it a correctness theorem, and this is precisely this main theorem which is going to help us build the tactic we want.	
$decideEq\_correct\ :\ \forall\  x\ y:c,\ decideEq\ x\ y\ =\ true\ \rightarrow\ x\ =\ y$. This proof can be made by combining, in this order, the three lemmas $beq_{Expr}\_correct$, $normalize\_correct$ (two times, one on each side of the assumed equality), and finally $reflect\_and\_reify\_correct$ (again two times).

Now, the tactic can be build by using this correctness theorem. When trying to prove a goal $a\ =\ b$, the tactic will apply this theorem $decideEq\_correct$, which means that $x$ will be unified to $a$ and $y$ to $b$. The desired proof is thus produced if the premise $decideEq\ a\ b\ =\ true$ holds. But whether this premise holds or not can simply be tested by running the function $decideEq$ on the two arguments $a$ and $b$.

@@

@@

Instead, what they did for Coq's implementation of the Ring prover is to replace the main theorem $\forall\  x\ y:c,\ decideEq\ x\ y\ =\ true\ \rightarrow\ x\ =\ y$ by the following weaker --but still powerful enough to build the desired tactic-- lemma\footnote{This lemma --that they call $setpolynomial\_simplify\_ok$ can be found in $plugins/ring/Setoid\_ring\_normalize.v$}  : $f\_correct\ :\ \forall\ (e1\ e2\ :\ Expr),\ beq_{Expr}\ (norm\ e1)\ (norm\ e2)\ =\ true\ \rightarrow\ reify\ e1\ =\ reify\ e2$.

@@

if it couldn't unify the goal $a=b$ with $reify\ (reflect\ a)\ =\ reify\ (reflect\ b)$, then it means that there is something wrong in the definitions of the functions $reflect$ and $reify$, and it could print something like "The ring prover has failed unexpectedly. There's something wrong with the implementations of the $reflect$ and $reify$ functions". Nothing really bad happened, as no inconsistent proof has been produced, nor no axioms added. That would just mean that the implementation of the ring prover is slightly broken, and that some goals that should be automatically provable aren't automatically proved. Thus, it would only decrease the completeness of the prover. That would be of course bad --because in the extreme case, the prover never succeed in a generating a proof and is thus completely useless--, but that would only limit the scope of usage of the prover.

@@

Completeness means that when given an equality which is provable by using the properties of the concerned algebraic structure, the prover should be able to generate a proof of this equality. Having a complete tactic is important, because otherwise the tactic is not always usuable -- and in the worst case it is never usable--, and thus is not valuable. However, having a formal proof of the completeness in the system itself isn't really interesting. As long as we do not meet an equality that should be provable, and which isn't proved by our tactic, there is no problem. And if this situation does happen, nothing really bad really happened : no inconsistencies have been introduced in the system. The tactic just refuses to do it automatically for the programmer. The user can still attempt it. \\

If we believe that the normal form computed by the reduction function is indeed a normal form --which means that if two terms are equal according to the properties of an algebraic structure, then their normal form should be equal-- then the tactic is complete.
Indeed, 
$meta\_propetry\ :\ \forall\ a\ b,\ a=b \rightarrow norm\ (reflect\ a) = norm\ (reflect\ b)$ is equivalent to

>>>>

Without our type-safe reflection mechanism, the traditional way to go for this problem would be to have functions doing only computations, and some external lemmas about them. It would start with the definition of the function $reify\ :\ Expr\ \rightarrow\ c$. As we did in our approach, there would still be a function for reflecting terms $reflect\ :\ c\ \rightarrow\ Expr$. Because $reflect$ is a function based on syntax,  the actual way to define it would depend on the facilities of the proof assistant / programming language used, but there need to be a way to write functions based on syntax. We would need to prove the correctness of these two functions, and it turns out that one lemma is enough for completely specifying the expected behaviour of these two functions at the same time : $reflect\_and\_reify\_correct\ :\ \forall\ x:c,\ reify\ (reflect\ x)\ =\ x$.

Then comes the normalisation function : $normalize\ :\ Expr\ \rightarrow\ Expr$. Because this function is weekly typed compared to our approach (we no longer have an index giving a guarantee about the concrete values), we would need to provide a lemma of correctness about it after its definition. This lemma of correctness must say that the interpretation of the original (reflected) term is equal to the interpretation of the normalised (reflected) term. $normalize\_correct\ :\ \forall\ e:Expr,\ reify\ (normalize\ e)\ =\ reify\ e$. We have avoided the need of such a --complicated-- proof in our correct by construction approach enabled by our type-safe reflection mechanism.

Now, similarly to what we've done, we would need a syntactical equality test, checking whether two reflected expressions are syntactically equal. The difference here is that, as often with this kind of "traditional" approach, we would produce something that belongs to the world of computations, usually a (pretty uninformative) boolean.
$beq_{Expr}\ :\ Expr\ \rightarrow\ Expr\ \rightarrow\ bool$. Because this boolean on its own is pretty uninformative, we now need a proof of correctness for this function as well, which says that if this function decides that two given terms are syntactically equal, then their interpretation should also be (syntactically) equal. $beq_{Expr}\_correct\ :\ \forall (e1\ e2:Expr), beq_{Expr}\ e1\ e2\ =\ true\ \rightarrow\ reify\ e1\ =\ reify\ e2$. We did not have to do this proof either with our approach.

We could now build the kernel of the tactic, which is the following decision procedure. 
\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
decideEq : c -> c -> bool 
decideEq x y =
      let ex = reflect x in 
      let ey = reflect y in
      beq_Expr (normalise ex) (normalise ey)
\end{verbatim}
\end{center}
\caption{Decision procedure of the tactic with a "traditional" approach}
\figrule
\end{figure}
This function on its own would not  be enough. We would need to add a correctness theorem, and this is precisely this main theorem that would help us to build the tactic we want.	
$decideEq\_correct\ :\ \forall\  x\ y:c,\ decideEq\ x\ y\ =\ true\ \rightarrow\ x\ =\ y$. This proof can can made by combining, in this order, the three lemmas $beq_{Expr}\_correct$, $normalize\_correct$ (two times, one on each side of the assumed equality), and finally $reflect\_and\_reify\_correct$ (again two times).

Now, the tactic could be built by using this correctness theorem. When trying to prove a goal $a\ =\ b$, the tactic would apply this theorem $decideEq\_correct$, which means that $x$ will be unified to $a$ and $y$ to $b$. The desired proof is thus produced if the premise $decideEq\ a\ b\ =\ true$ holds. But whether this premise holds or not can simply be tested by running the function $decideEq$ on the two arguments $a$ and $b$.

@@

		\subsubsection{Description of Coq's implementation}

@@

Instead, what they did for Coq's implementation of the Ring prover was to replace the main theorem $\forall\  x\ y:c,\ decideEq\ x\ y\ =\ true\ \rightarrow\ x\ =\ y$ by the following weaker --but still powerful enough to build the desired tactic-- lemma\footnote{This lemma --that they call $setpolynomial\_simplify\_ok$-- can be found in $plugins/ring/Setoid\_ring\_normalize.v$}  : $f\_correct\ :\ \forall\ (e1\ e2\ :\ Expr),\ beq_{Expr}\ (norm\ e1)\ (norm\ e2)\ =\ true\ \rightarrow\ reify\ e1\ =\ reify\ e2$.

@@

if it couldn't unify the goal $a=b$ with $reify\ (reflect\ a)\ =\ reify\ (reflect\ b)$, then it means that there is something wrong in the definitions of the functions $reflect$ and $reify$, and it could print something like "The ring prover has failed unexpectedly. There's something wrong with the implementations of the $reflect$ and $reify$ functions". Nothing really bad happened, as no inconsistent proof has been produced, nor no axioms added. That would just mean that the implementation of the ring prover is slightly broken, and that some goals that should be automatically provable aren't automatically proved. Thus, it would only decrease the completeness of the prover. That would be of course bad --because in the extreme case, the prover never succeeds in a generating a proof and is thus completely useless--, but that would only limit the scope of usage of the prover. \\

		\subsubsection{Differences with Coq's implementation}

Coq's ring prover follows the traditional approach of defining functions, and latter on proving many auxiliary lemmas about them. This is particularly adapted to Coq, which has many facilities for the construction of proofs, and an adapted proof mode. However, this approach kind of duplicate the work : they first tell the machine how it works, and latter, why it works. This separation between the world of computations and the world of logic becomes smaller in our approach with a good use of dependent types, that allows to write more specific types, and thus to capture logical properties. The writing of functions can therefore be guided by this more precise type, and this is one of the main benefit of the approach we've followed. Moreover, we almost get the proof of correctness for free, because every little bit of rewriting done for the normalisation of the reflected terms is accompanied by the logic justification which tells why this rewriting can be done --and locally, this justification is always simple !--. In the end, all the proofs we have to do are systematically trivial, since they only contain trivial rewritings with the available hypothesis and the use of the properties of the corresponding algebraic structure. \\

Another difference with our implementation is that we've implemented a hierachy of tactics for many algebraic structures, but Coq's implementation only deals with ring and semi ring. If someone wants to prove equalities in a commutative group that isn't a ring, he simply can't use their prover. A dedicated prover for commutative groups would be needed, and Coq currently doesn't has one. The worst is that such a prover for commutative group would do very similar treatments, which means a lot of code could have been factorised. This is what we've obtained by taking this in account from the start. Our monoid prover used the underneath semi-group prover, our group prover uses the underneath monoid prover, and so on. \\

[Mentioning performances, that might be in favour of Coq]

@@

Completeness means that when given an equality which is provable by using the properties of the concerned algebraic structure, the adapted prover should be able to generate a proof of this equality. Having a complete tactic is important, because otherwise the tactic is not always usuable -- and in the worst case it is never usable--, and thus is not really valuable. However, having a formal proof of the completeness in the system itself isn't really interesting. As long as we do not meet an equality that should be provable, and which isn't proved by our tactic, there is no problem. And if this situation does happen, nothing really bad really happened : no inconsistencies have been introduced in the system. The tactic just refuses to do it automatically for the programmer. The user can still attempt it. \\

If we believe that the normal form computed by the reduction function is indeed a normal form --which means that if two terms are equal according to the properties of an algebraic structure, then their normal form should be equal-- then the tactic is complete.
Indeed, 
$meta\_propetry\ :\ \forall\ a\ b,\ a=b \rightarrow norm\ (reflect\ a) = norm\ (reflect\ b)$ is logically equivalent to

### 2016-01-12 paper-JFP-DTP/paper.tex
paper...
paper JFP DTP continued
Renamed content files
paper continued and corrected. Still roughly 2 pages to write

\begin{abstract}
% The abstract should be at most 300 words for the Journal of Functional Programming
Idris is a general purpose purely functional programming language with dependent types, aiming to bring type-based program verification techniques to functional programmers. One common difficulty with programming with dependent types is that proof obligations arise naturally once programs become even moderately sized. For example, implementing an adder for binary numbers indexed over their natural number equivalents will naturally lead to proof obligations for equalities of expressions over natural numbers. 
As far as possible, we would like to solve such proof obligations automatically. In this paper, we show one way to automate such proofs by reflection. We will show how representing Idris expressions in a reflected form, indexed by the original Idris expression, leads to straightforward construction and manipulation of proofs.
With this type-safe reflection technique, the resulting expressions are guaranteed to be faithful representations of the corresponding inputs and any generated proof is guaranteed to be a proof of the required property. 
We will first present the technique on a small example, where we aim to automatically prove equalities with universally quantified lists and the associative concatenation operation, and later show how this idea is generalised to various kind of properties (or axioms) that might be available (for example : commutativity, distributivity, existence of neutral elements...), leading to a hierarchy of reflexive tactics for Monoids, Commutative Monoids, Groups, Commutative Groups, Rings, and so on, written in Idris, for proving different kind of equivalence. We will also show how each tactic reuses the other ones from the simplest structures, thus avoiding as much as possible the duplication of code.
\end{abstract}

@@

\input{./content/introduction.tex}

\input{./content/basicIdeas.tex}

\input{./content/generalProblem.tex}

\input{./content/automaticReflection.tex}

\input{./content/discussion.tex}

@@

>>>>

\begin{abstract}
% The abstract should be at most 300 words for the Journal of Functional Programming
Idris is a general purpose purely functional programming language with dependent types, aiming to bring type-based program verification techniques to functional programmers. One common difficulty with programming with dependent types is that proof obligations arise naturally once programs become even moderately sized. For example, implementing an adder for binary numbers indexed over their natural number equivalents will naturally lead to proof obligations for equalities of expressions over natural numbers. 
As far as possible, we would like to solve such proof obligations automatically. In this paper, we show one way to automate such proofs by reflection. We will show how representing Idris expressions in a reflected form, indexed by the original Idris expression, leads to straightforward construction and manipulation of proofs.
With this type-safe reflection technique, the resulting expressions are guaranteed to be faithful representations of the corresponding inputs and any generated proof is guaranteed to be a proof of the required property. 
We will first present the technique on a small example, where we aim to automatically prove equalities with universally quantified natural numbers and the associative addition operation, and later show how this idea can be generalised to various kind of properties (or axioms) that might be available (for example : commutativity, distributivity, existence of neutral elements...), leading to a hierarchy of reflexive tactics for Monoids, Commutative Monoids, Groups, Commutative Groups, Rings, and so on, written in Idris, for proving different kind of equivalence. We will also show how each tactic reuses the other ones from the simplest structures, thus avoiding as much as possible the duplication of code.
\end{abstract}

@@

\input{./content/1-introduction.tex}

\input{./content/2-basicIdeas.tex}

\input{./content/3-generalProblem.tex}

\input{./content/4-detailsNormalisation.tex}

\input{./content/5-discussion.tex}

@@

 \bibitem[\protect\citename{W.A Howard}1980]{How80}
   Howard W.A. The formulae-as-types notion of construction. In J. R. Seldin and J. P. Hindley, editors, To H. B. Curry: Essays on Combinaory Logic, Lambda Calculus, and Formalism. Academic Press, 1980.
